---
title: "Practical Machine Learning Course Project"
output: html_document
---
#####Nick Franciose

###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Load the data
```{r}
library(AppliedPredictiveModeling)
library(caret)
library(rpart)
library(rattle)
library(rpart.plot)
library(randomForest)
library(knitr)
library(RColorBrewer)
library(ggplot2)
library(plyr)
#Read files and replace zeros with NA
train <- read.csv("pml-training.csv", na.strings=c("NA","#DIV/0!", ""), header=TRUE)
test <- read.csv("pml-testing.csv", na.strings=c("NA","#DIV/0!", ""), header=TRUE)
```

####Clean the data


```{r}
#Remove the first seven columns
train <- train[-c(1:7)]

#Remove NearZeroVariance variables

#nearZeroVar <- nearZeroVar(train, saveMetrics=TRUE)
#train <- train[!nearZeroVar]


#Clean variables with more than 70% NA

cleanedtrain<- train
for(i in 1:length(train)) {
    if( sum( is.na( train[, i] ) ) /nrow(train) >= .7) {
        for(j in 1:length(cleanedtrain)) {
            if( length( grep(names(train[i]), names(cleanedtrain)[j]) ) == 1)  {
                cleanedtrain <- cleanedtrain[ , -j]
            }   
        } 
    }
}

# Set back to the original variable name
train <- cleanedtrain
rm(cleanedtrain); rm(i); rm(j)

#Explore the Test Data
ggplot(train, aes(total_accel_belt, colour = classe, fill = classe)) +geom_density(alpha = 0.1) 
ggplot(train, aes(total_accel_arm, colour = classe, fill = classe)) +geom_density(alpha = 0.1)
ggplot(train, aes(total_accel_dumbbell, colour = classe, fill = classe)) +geom_density(alpha = 0.1)
ggplot(train, aes(total_accel_forearm, colour = classe, fill = classe)) +geom_density(alpha = 0.1)
#Partioning the training set into two
trainPart <- createDataPartition(train$classe, p=0.6, list=FALSE)
myTrain <- train[trainPart, ]
myTest <- train[-trainPart, ]
dim(myTrain); dim(myTest)

#Classification Tree

set.seed(666)
ctFit <- train(myTrain$classe ~ ., data = myTrain, method="rpart")
print(ctFit, digits=3)

fancyRpartPlot(ctFit$finalModel)

#Apply Classification tree to myTest
ctPredictions <- predict(ctFit, newdata=myTest)
print(confusionMatrix(ctPredictions, myTest$classe), digits=4)

#Create a prediction model using boosting with Trees

#TRAIN
gbmFit <- train(classe ~ ., data = train, method = "gbm", verbose = FALSE)

#Test Boosting with Trees model
gbmPrediction <- predict(gbmFit, myTest)
confusionMatrix(gbmPrediction, myTest$classe)


#Create Random Forest model
set.seed(1777)
random_forest=randomForest(classe~.,data=myTrain,ntree=500,importance=TRUE)
random_forest

plot(random_forest,main="Error Rate vs Number of Trees")

rfPredictions = predict(random_forest, newdata=myTest)
confusionMatrix(rfPredictions,myTest$classe)


#Explore Variable importance to Random Forest model


```

#Applying RF model to test set
The random forest is the most accurate, so we will apply this model to the test set. 

```{r}
#Remove all unneeded columns
test <-test[,c(8:11,37:49,60:68,84:86,102,113:124,140,151:160)]
print(predict(random_forest, newdata=test))
```


